Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-16 manuscript.tex
edits to abstract

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the \emph{Hopscotch} element, led us to investigate the prevalence and potential role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of both subspecies \emph{Zea mays} ssp. \emph{parviglumis} and subspecies \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments fail to find an important role for \emph{tb1} in controlling tillering in natural populations of \emph{parviglumis}. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte is not as straightforward as is in domesticated maize, and that other loci may play a role in observed variation in tillering in teosinte.

>>>>

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the \emph{Hopscotch} element, led us to investigate the prevalence and potential role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest an important role for \emph{tb1} in controlling tillering in natural populations of this subspecies. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte is not as straightforward as is in domesticated maize, and that other loci may play a role in observed variation in this trait.
